---
title: "Lasso screening"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
link-citations: TRUE
vignette: >
  %\VignetteIndexEntry{Lasso screening}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.5
)
# The simulations below take several minutes. By default, the vignette loads
# precomputed summaries from vignettes/data/. Set the environment variable
# SIMCITY_RUN_SIMS=true to rerun them and overwrite those files.
run_sims <- identical(Sys.getenv("SIMCITY_RUN_SIMS"), "true")
```

## Lasso variable selection

In high-dimensional regression with $n$ observations, $p$ predictor variables, and $p > n$, we cannot fit ordinary least squares. One strategy around this is to **bet on sparsity**: we assume there is a true (or good enough) linear model with a relatively small number $s_0 < n$ of nonzero coefficients.

The lasso [-@tibshirani1996] can be used to fit a linear model to high-dimensional data. It typically results in a fit with few nonzero coefficient estimates. The `glmnet` [package](https://cran.r-project.org/package=glmnet) [-@glmnet] is a standard software implementation that can also use cross-validation to choose the regularization parameter.

In this vignette we show how to use `simcity` to simulate many high-dimensional datasets where the true model is a sparse linear model. Designs and coefficients are generated with `hdi::rXb()` [@rhdi]. We fit the lasso with `glmnet::cv.glmnet` at `lambda.1se`, and then check whether the estimated model has a property we call **screening**. Writing $S = \{j : \beta_j \neq 0\}$ and $\hat S = \{j : \hat\beta_j \neq 0\}$, screening means $S \subseteq \hat S$: the selected variables include all the ones with nonzero coefficients in the true model.

```{r setup}
library(ggplot2)
theme_set(theme_minimal())
library(simcity)
```

### A single simulated dataset

First let's look at one example instance, showing both the true and estimated coefficients:

```{r}
set.seed(1)
n <- 50
p <- 100
s0 <- 2
one_lasso_fit <- instance_hdr(n, p, s0)
head(one_lasso_fit)
```

The first row is the intercept. Its true value is 0, and it is not penalized or counted as a selected variable. Dropping it, which coefficients are nonzero in the true model? `hdi::rXb()` always puts the true support first. With the default `permuted = TRUE`, it instead shuffles the columns of $X$, so the active variables are not neighbours in the Toeplitz correlation structure.

```{r}
slopes <- one_lasso_fit[one_lasso_fit$term != "(Intercept)", ]
true_support <- which(slopes$true_beta != 0)
true_support
```

Which coefficients are nonzero in the lasso selected model?

```{r}
estimated_support <- which(slopes$estimate != 0)
estimated_support
```

Did the selected model succeed at screening?

```{r}
all(true_support %in% estimated_support)
```

### Many simulated datasets

We'll use `simulate_hdr()`, which runs many instances in parallel with reproducible random number streams. `simmary_coefs()` then computes support recovery metrics for each instance.

```{r}
n <- 100
p <- 200
s0 <- 5
niters <- 100
```

```{r, eval = run_sims}
many_lasso_fits <- simulate_hdr(niters, n, p, s0, seed = 1)
sim_summary <- simmary_coefs(many_lasso_fits)
write.csv(sim_summary, "data/lasso-screening.csv", row.names = FALSE)
```

```{r}
sim_summary <- read.csv("data/lasso-screening.csv")
head(sim_summary)
```

What was the frequency of screening? With `r niters` independent replications, a binomial confidence interval conveys the Monte Carlo error:

```{r}
k <- sum(sim_summary$screened)
k / nrow(sim_summary)
binom.test(k, nrow(sim_summary))$conf.int
```

Was the smallest truly nonzero coefficient different in the cases when screening occurred?

```{r}
ggplot(sim_summary, aes(screened, beta_min)) +
  geom_boxplot()
```

### Sparsity and screening probability

Let's now compute the rate of screening as a function of the sparsity of the true model, keeping all other parameters fixed. The helper below collapses the per-instance summaries into one row per sparsity level. It adds a Clopper–Pearson 95% interval for the screening probability and the mean false discovery proportion.

```{r}
summarize_by_sparsity <- function(full_results) {
  by_s0 <- split(full_results, full_results$true_sparsity)
  do.call(rbind, lapply(by_s0, function(d) {
    k <- sum(d$screened)
    ci <- binom.test(k, nrow(d))$conf.int
    data.frame(true_sparsity = d$true_sparsity[1],
               n_sims = nrow(d),
               screening = k / nrow(d),
               lower = ci[1],
               upper = ci[2],
               mean_fdp = mean(d$fdp),
               avg_bmin = mean(d$beta_min))
  }))
}

run_sparsity_sims <- function(sparsities, niters, ...) {
  full_results <- do.call(rbind, lapply(sparsities, function(s0) {
    simmary_coefs(simulate_hdr(niters, n, p, s0, ..., seed = s0))
  }))
  summarize_by_sparsity(full_results)
}
```

```{r}
n <- 100
p <- 200
sparsities <- 1 + 2 * 0:15
niters <- 200
```

```{r, eval = run_sims}
sim_summary <- run_sparsity_sims(sparsities, niters, verbose = TRUE)
write.csv(sim_summary, "data/lasso-screening-sparsity.csv", row.names = FALSE)
```

```{r sparsity-plot1}
sim_summary <- read.csv("data/lasso-screening-sparsity.csv")
ggplot(sim_summary, aes(true_sparsity, screening)) +
  geom_col(aes(fill = avg_bmin)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.4) +
  scale_fill_viridis_c(direction = -1)
```

The frequency of screening decreases as the true sparsity increases. But the true coefficients are, by default, drawn from a uniform distribution $U[-2, 2]$. So the average (over simulation instances) minimum coefficient, `avg_bmin`, also decreases with $s_0$: the minimum of $s_0$ i.i.d. $|U[-2,2]|$ draws has expectation $2/(s_0+1)$.

To separate these effects, let's run the simulation again with all nonzero coefficients fixed at 1, using `btype = "bfix1"`.

```{r}
sparsities <- 9 + 2 * 0:16
```

```{r, eval = run_sims}
sim_bfix_summary <- run_sparsity_sims(sparsities, niters, btype = "bfix1",
                                      verbose = TRUE)
write.csv(sim_bfix_summary, "data/lasso-screening-sparsity-bfix1.csv",
          row.names = FALSE)
```

```{r sparsity-plot2}
sim_bfix_summary <- read.csv("data/lasso-screening-sparsity-bfix1.csv")
ggplot(sim_bfix_summary, aes(true_sparsity, screening)) +
  geom_col(fill = "grey60") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.4)
```

With every nonzero coefficient equal to 1, $\beta_{\min}$ no longer shrinks as $s_0$ grows. Screening remains likely at sparsity levels where it had become rare with uniform coefficients. For example, at $s_0 = 21$ the rate is about 95% here, compared with about 5% above. It still breaks down once $s_0$ is large relative to $n$. This is qualitatively consistent with theory for lasso support recovery, which requires a sample size of order $s_0 \log(p - s_0)$ [@wainwright2009]. Here $n / \log p \approx 100 / 5.3 \approx 19$, which matches where screening starts to decline, but only up to constants. That theory concerns exact recovery with a deterministic $\lambda$, whereas screening with a cross-validated $\lambda$ is a weaker property, so the comparison is only heuristic.

### Other metrics

Lasso screening is just one example. This package can be used to compare various performance metrics, fitting procedures, and other kinds of simulated data and tasks. As a brief example, here is one other metric from the simulations above: the **false discovery proportion** (FDP). The FDP is the fraction of selected variables whose true coefficient is zero, $|\hat S \setminus S| / \max(1, |\hat S|)$, and its mean over instances estimates the false discovery rate.

```{r sparsity-plot-fdp}
ggplot(sim_bfix_summary, aes(true_sparsity, mean_fdp)) +
  geom_col(aes(fill = screening)) +
  scale_fill_viridis_c()
```

## References
