---
title: "Parallel computation and reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel computation and reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simcity)
```

`simulate_hdr()` runs independent replications of `instance_hdr()` with
`foreach` and `doRNG`. With `cores > 1` it starts a local PSOCK cluster,
which is stopped again when the function returns.

## Timing

The timings below are from whatever machine built this page, with at most
2 cores so that the vignette also builds on CRAN-like systems. Treat them as
an illustration, not a benchmark. Starting worker processes has a fixed cost
of roughly a second, so parallelism pays off only when each replication or
the number of replications is large enough.

```{r}
n <- 100
p <- 200
s0 <- 10
niters <- 40
```

Sequentially, with `replicate()`:

```{r replicate}
set.seed(1)
system.time(
  many_lassos_unparallel <- replicate(niters, instance_hdr(n, p, s0),
                                      simplify = FALSE)
)
```

With `simulate_hdr()` on one core (sequential, in this R session) and on two
cores:

```{r cores}
system.time(
  many_lassos_1core <- simulate_hdr(niters, n, p, s0, cores = 1, seed = 1)
)
system.time(
  many_lassos_2cores <- simulate_hdr(niters, n, p, s0, cores = 2, seed = 1)
)
```

On a machine with more cores, `simulate_hdr()` defaults to half of
`parallel::detectCores()`.

## Reproducibility

Each replication gets its own L'Ecuyer-CMRG random number stream, derived from
`seed`. So the results depend on `seed` but not on the number of cores:

```{r}
identical(many_lassos_1core, many_lassos_2cores)
```

The seed of each replication is stored in the `"rng"` attribute. This lets
you rerun a single replication, for example to investigate an unusual
result:

```{r}
rng_states <- attr(many_lassos_2cores, "rng")
assign(".Random.seed", rng_states[[7]], envir = globalenv())
rerun_7 <- instance_hdr(n, p, s0)
all.equal(rerun_7, many_lassos_2cores[[7]], check.attributes = FALSE)
RNGkind("default")
```

## Custom components on workers

Workers are fresh R processes. Functions you pass as `yfun`, `fitfun` or
`postfun` are sent to them along with the objects they refer to. Functions
from other packages should be called with `pkg::fun()`, or the packages
listed in the `packages` argument, so that they are loaded on each worker.
